---
title: "datacolor"
author: Éric Marty
date: August 1, 2018
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

Color for Data Visualization
<!-- https://ecotrust-canada.github.io/markdown-toc/ -->

# Table of Contents

- [Introduction](#introduction)
  * [Installation](#installation)
- [Analysis and display of palettes](#analysis-and-display-of-palettes)
- [Converters](#converters)
  * [Hex RGB to HCL](#hex-rgb-to-hcl)
  * [Hex RGB palette to simulated colorblind hex RGB](#hex-rgb-palette-to-simulated-colorblind-hex-rgb)
- [Precalculated palettes](#precalculated-palettes)
  * [7 category qualitative palette for nominal data](#7-category-qualitative-palette-for-nominal-data)
  * [Sequential palette for heatmaps](#sequential-palette-for-heatmaps)
    + [Comparison of Heat Palettes](#comparison-of-heat-palettes)
  * [Divergent colorscale for AUC heatmaps](#divergent-colorscale-for-auc-heatmaps)
- [HCL Palettes with power curves, stepped and cyclic elements](#hcl-palettes-with-power-curves-stepped-and-cyclic-elements)
- [Multihue HCL Palettes with bezier interpolation and lightness correction](#multihue-hcl-palettes-with-bezier-interpolation-and-lightness-correction)
  * [Comparison with non-bezier multi-hue palettes](#comparison-with-non-bezier-multi-hue-palettes)

# Introduction

This package provides  

* prebuilt color palettes appropriate for data representation,  
* functions for defining color palettes for data representation, 
* utilities for analysis of palettes as perceived by normal and colorblind viewers.


Color palettes for data should ideally be perceptually linear in all perceptual color dimensions 
(hue, chroma or saturation, lightness or luminance). Ideally, palettes should aslo be color-blind 
safe (i.e. they should preserve an appropriate mapping of data to perceived color for colorblind 
viewers). Palettes may also need to be black and white printer-friendly, if there is a chance plots 
will be printed in greyscale. Additional requirements for color scales can include the presence or 
absence of discontinuities and corners, maximizing discrimination between levels, and respecting 
semantic relationships between colors.  

This package relies on the perceptually flat colorspace HCL, which represents color as preceived 
by people with normal vision. HCL functions from the `grDevices` and `colorspace` packages. This 
package also relies on the `dichromat` R package to simulate color perception under the three most 
common forms of colorblindness.  

## Installation

Install the `datacolor` package from github:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
# devtools::install_github("allopole/datacolor")
```

# Analysis and display of palettes

See `?colorbar` for description of `colorbar` and `colorplot` functions

```{r colorbar}
library(datacolor)

# colorbar
pal <- grDevices::rainbow(36)
colorbar(pal)
colorbar(pal,dots=TRUE)
colorbar(pal,colorblind=TRUE)
colorbar(pal,dots=TRUE,colorblind=TRUE)
```

```{r colorplot}
# colorplot
colorplot(pal)
colorplot(pal,dots=TRUE)
colorplot(pal,colorblind=TRUE)
colorplot(pal,dots=TRUE,colorblind=TRUE)
```


# Converters

## Hex RGB to HCL

See `?hex2hcl`

```{r hex2hcl}
pal <- grDevices::rainbow(7)
hex2hcl(pal)
```

## Hex RGB palette to simulated colorblind hex RGB

See `?colorblind`

```{r colorblind}
pal <- grDevices::rainbow(7)
colorblind(pal)
colorplot(colorblind(pal)$deuteranopia)
```

# Precalculated palettes

## 7 category qualitative palette for nominal data

See `?unipalette` for details

* designed for high contrast against a white background, with good distinction among colors
* colors distinguishable under two most common forms of colorblindness
* colors distinguishable in greyscale for b/w printing
* Good for bar charts, box plots, pies, etc.
* Line should be relatively thick
* Points should be relatively large


```{r unipalette}
colorplot(unipalette(),colorblind = TRUE)

## Sample plot
randomseries <- data.frame(x=1:4, y=runif(4*7), grp=rep(letters[1:7],each=4))
pal <- unipalette()
lattice::xyplot(y~x, group=grp, data=randomseries, type="b", pch=16, lwd=3, col = pal,
      key = list(space = "right", lines = list(col=pal, lwd=3), text=list(letters[1:7]))
      )
```

## Sequential palette for heatmaps

A replacement for `grDevices::heat.colors()` or `colorspace::heat_hcl()`

See `?heat` for details

Palette characteristics:

* perceptually linear
* color-blind-safe (under three most common forms of colorblindness) and print friendly
* under normal vision: linear lightness, chroma, and hue gradients
* under colorblind viewing, linear lightness, smooth chroma, flat hue gradients
* appropriate for display of temperature, intensity, etc.

```{r heat}
colorplot(datacolor::heat(128),colorblind = T)
mtext("datacolor::heat()", side=1, line = 2.5)

```

### Comparison of Heat Palettes

```{r heat_comparison}
n <- 128

colorplot(grDevices::heat.colors(n), colorblind = TRUE)
mtext("grDevices::heat.colors(128)", side=1, line = 2.5)

colorplot(viridis::inferno(128), colorblind = TRUE)
mtext("viridis::inferno(128)", side=1, line = 2.5)

colorplot(colorspace::heat_hcl(n), colorblind = TRUE)
mtext("colorspace::heat_hcl()", side=1, line = 2.5)

colorplot(datacolor::heat(128),colorblind = T)
mtext("datacolor::heat()", side=1, line = 2.5)

## Maroon to yellow through salmon and pink
stops <- c('#800000', '#CD4949', '#FF9696','#FFFF00')
palette <- multiHue(128,colors=stops)
colorplot(palette,colorblind = TRUE) # display palette
mtext("datacolor::multiHue(128,c('#800000', '#CD4949', '#FF9696','#FFFF00'))", side=1, line = 2.5)


```

## Divergent colorscale for AUC heatmaps

See `?AUColors` for details

Palette characteristics:

* color-blind-safe (under three most common forms of colorblindness)  
* perceptually linear  
* equal brightness gradients on both sides of central value  
* constant hue and chroma on either side of central value  
* Dark value for no correlation (near 0.5); bright values for high correlation (near 1 and 0)  
* Blue for positive correlation (1); orange for negative correlation (0)  
* NOT suitable for b/w printing, as low and high values are not distinguishable in greyscale (however, greyscale printing still accurately shows the absolute distance from 0.5)  

```{r AUC}
## default palette (n = 10):
p <- AUColors()
colorplot(p,colorblind = TRUE)

## inverted and reversed, with grey center bin:
p <- AUColors(n=21,invert=TRUE,reverse=TRUE)
colorplot(p,colorblind = TRUE)

# Sample AUC Heat Map
randommatrix <- matrix(runif(12*6),ncol=6)
nlevels <- 10 
lattice::levelplot(randommatrix,
  at = seq(0,1,1/nlevels), # number of breaks
  col.regions = AUColors(nlevels), # color map
  aspect = "iso")


```


# HCL Palettes with power curves, stepped and cyclic elements

Use `rampx()`, `stepx()`, and `cyclx()` to construct numeric vectors of `H`, `C`, and `L`. 

See `?cyclx` for description of `rampx()`, `stepx()`, and `cyclx()`.

Use `hcl2hex()` to convert HCL palette to a hex RGB palette.

Depricated: `cyclic_hcl`

```{r cyclx}
library(datacolor)

n <- 36

pal <- hcl2hex(
  H=rampx(from=260,to=-100,n=n),
  C=stepx(from=40,to=60,n=n,step.n=n/3),
  L=100*cyclx(from=.75,to=.45,n=n,
              step.n=n/3,exponent=1,
              cyc.from=.8,cyc.to=1,cyc.n=n/3,cyc.exponent=2.5)
  )

colorplot(pal)
```



# Multihue HCL Palettes with bezier interpolation and lightness correction

`multihue()`, `multihue.diverge()` and `multihue.constantL()` create multi-hue (multi "stop")
color scales with perceptually linear lightness gradients and smooth hue gradients (with no
first or second order discontinuities). 

The palettes follow bezier curves through HCL space. The bezier curves are defined by two endpoint 
colors and one or more optional intermediate color "stops."  The intermediate stops are treated as 
control points for the bezier curve, and are therefore not necessarily present in the final palette. 

After interpolation, the palettes are stretched to counter any non-linearity in lightness gradient 
arising from the presence of intermediate color stops. The resulting palette has a perceptually 
linear lightness gradient.

`multihue.diverge()` builds a divergent palette by concatenating two bezier curves (defined by a 
"left" and "right" set of color stops).

See `?multiHue` for description of `multiHue()`, `multiHue.diverge()` and `multiHue.constantL()`

For a discussion of bezier interpolation in HCL with lightness correction, see:

Gregor Aisch. *Mastering Multi-hued Color Scales with Chroma.js.* Blog post. Sep 9, 2013 <https://www.vis4.net/blog/2013/09/mastering-multi-hued-color-scales/>

Palette characteristics:

* perceptually linear lightness gradients (or constant lightness for `multihue.constantL()` )
* smooth hue and chroma gradient throughout (no "corners")
* multiple color stops allow increased color discrimination between levels
* hue and chroma gradients not necessarily linear. Only brightness is adjusted for linearity.
* `multihue()`, `multihue.diverge()` produce print-friendly palettes.
* `multihue.diverge()` may produce a palette with a central discontinuity (first or second order)


```{r multiHue}
## default = blue to yellow through green, 11 colors
p <- multiHue()
colorplot(p,colorblind=TRUE)
colorplot(p)

## Maroon to yellow through salmon and pink
stops <- c('#800000', '#CD4949', '#FF9696','#FFFF00')
palette <- multiHue(128,colors=stops)
colorplot(palette, colorblind=TRUE) # display palette

## Divergent (default)
p <- multiHue.diverge(128)
colorplot(p,colorblind=TRUE)

## Divergent with discontinuity: left = Blue to Cyan, right = Yellow to Red
leftstops <- c("#3136AA","#3980B0","#00C5C0","#8EFDFD")
rightstops <- c("#F3F300","#FF8CB4","#CD4C4C","#8F0000")
palette <- multiHue.diverge(128,leftstops,rightstops,continuous=FALSE)
colorplot(palette,colorblind=TRUE)

## Constant Lightness (and chroma) color palette:
## 4 color stops define the bezier curve through HCL space; n=6 colors are output.
palette <- multiHue.constantL(n=64, h=c(-180, -105, -45, 0),c=50,l=60)
colorplot(palette,colorblind=TRUE)
```


## Comparison with non-bezier multi-hue palettes

By comparison, here is the "rainbow" palette showing discontinuities typical of multi-hule palettes. 
This palette includes corners (second order discontinuities) in lightness. From a perceptual point of view, this palette is not a continuous colors scale, but a multiply-diverging scale with three distinct articulation points (at 1/6, 1/2 and 5/6 of the palette length).

```{r rainbow}
pal <- grDevices::rainbow(128)
colorplot(pal,colorblind=TRUE)
```

